What Hast Earth to do With Mars: An explanation of Urbit's IPC interfaces

### 2024-01-26 mss.tex
added sections

\section{Background and Literature}

Exposite the relevant background and literature.

\begin{enumerate}
  \item  Prefer to \texttt{enumerate}.
  \item  It is easier to refer to than \texttt{itemize}.
\end{enumerate}

\noindent
Mark \texttt{\textbackslash noindent} in paragraphs that continue the thought of an \texttt{\textbackslash enumerate} block.

\section{Urbit's Implementation}

Oftentimes you will then turn to exploring how Urbit or closely related systems like Nockchain have approached or considered a problem.

\lstset{language=C}
\begin{lstlisting}
#include <stdio.h>

double compute(float a, float b) {
    return a * a + b * b;
}

int main() {
    float x = 2.5;
    float y = 3.7;

    double result = compute(x, y);

    printf("Result: %lf\n", result);

    return 0;
}
\end{lstlisting}

\noindent
That code snippet was part of the text, and not a standalone entity to which we make separate repeated reference.

\begin{figure}
  \begin{lstlisting}[language=Python, caption={Example Python Code}, label={lst:example}]
  def hello_world():
      print("Hello, world!")
  \end{lstlisting}
\end{figure}

Refer to the code listing: Listing~\ref{lst:example}.  (Note that we don't suppress indentation since it's a float.)

\section{A Wild Ernest Appears}

You can have as many sections as make sense.  Only sections and subsections appear in the table of contents.

\section{Conclusion}

>>>>

\section{Khan - The High-Level Thread Interface of Urbit}

Khan, as a crucial component of Urbit's architecture, functions as a high-level thread interface, pivotal for managing both internal and external communications within the Urbit ecosystem. It is designed to efficiently handle complex I/O operations, employing an IO monad coupled with an exception monad. This dual-mechanism approach allows Khan to adeptly manage complex IO tasks while effectively handling potential failures.

Khan's role in Urbit can be likened to that of a control plane. Its primary function is to execute threads via a Unix Socket and relay the results back. This setup supports various thread-running modes, enhancing the flexibility of Urbit's computing environment and allowing for diverse interaction methods with both internal and external processes.

At its core, Khan's design revolves around the use of 'threads', which are monadic functions designed to handle arguments and produce results. These threads are optimized for I/O operations, leveraging the IO monad for structured and efficient task handling. The exception monad further ensures robust management of potential failures and unexpected events.

Khan's implementation ensures that Urbit serves as a personal server, replacing multiple developer-hosted web services on foreign servers with self-hosted applications on a single personal server. Its design underscores Urbit's commitment to user ownership and data management, setting it apart from the conventional model of cloud-based services and centralized data management prevalent in Earth-based systems.

\section{\texttt{conn.c}} %Rewrite so its not a blatant copy of alex's work

\texttt{conn.c}\footnote{\url{https://github.com/urbit/vere/blob/develop/pkg/vere/io/conn.c}} is a driver in Vere. It is a part of the "King" (a.k.a. "Urth") process. It exposes a \texttt{Unix domain socket}\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}} at \texttt{/path/to/pier/.urb/conn.sock} for sending/receiving data from external processes.

The functionality of conn.c, particularly its ability to dispatch messages and handle various types of requests, is critical for the command of urbit's control plane. It serves as a way for unix processes to receive insights about what is happening on urbit. It also allows for pass commands either \%khan and the injection of raw kernel moves into the system. The implication of this means that whoever is running your urbit runtime controls your urbit.

From a technical perspective, conn.c accepts newt-encoded \texttt{++jam}med nouns which take the form \texttt{[request-id command arguments]}. This structure allows for a variety of commands to be executed, including

\begin{enumerate}
  \item \texttt{\%ovum} - the injection of raw kernel moves 
  \item \texttt{\%fyrd} - a direct shortcut to Khan commands
  \item \texttt{\%urth} - runtime subcommands like \texttt{\%pack}` or \texttt{\%meld}
  \item \texttt{\%peek} - namespace scry requests into Arvo
  \item \texttt{\%peel} - emulated namespace scry requests into Vere
\end{enumerate}

A valid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[request-id output]}, where:
\begin{itemize}
  \item \texttt{request-id} matches the input \texttt{request-id}
  \item \texttt{output} depends on the \texttt{command}
\end{itemize}

An invalid \texttt{conn.c} command produces a newt-encoded jammed noun with type \texttt{[0 \%bail error-code error-string]}.

\subsection{\%ovum}

The argument to an \texttt{\%ovum} command is a raw kernel move which is injected directly into the Arvo event loop. This is a very powerful - and potentially dangerous - tool. For example, if a ship somehow got into a state where Clay was no longer working properly (meaning new files could not be compiled to fix the state of the kernel), the source code for a new, working Clay could be directly injected into the ship using an \texttt{\%ovum}.

The output of an \texttt{\%ovum} command is:
\begin{itemize}
  \item \texttt{[\%news \%done]} if the move completed successfully
  \item \texttt{[\%news \%drop]} if the move was dropped
  \item \texttt{[\%bail goof]} if an error occurred
\end{itemize}

\subsection{\%fyrd}

\texttt{\%fyrd} is a direct shortcut to the Khan vane. The arguments to a \texttt{\%fyrd} command are (in order):
\begin{enumerate}
  \item The name of the desk in which the thread lives (e.g. \texttt{\%base}) or \texttt{beak} for the thread (e.g. \texttt{[\%zod \%base \%10]})
  \item The name of the thread (e.g. \texttt{\%hi})
  \item Mark to which the output should be cast (e.g. \texttt{\%tape})
  \item Mark for how to interpret the input argument to thread (e.g. \texttt{\%ship})
  \item Input argument to thread (e.g. \texttt{\~zod})
\end{enumerate}

The output of a \texttt{\%fyrd} command is \texttt{[\%avow (each page goof)]}, the value of \texttt{each} depending on whether the thread succeeded or not.

\subsection{\%urth}

The argument to the \texttt{\%urth} command is a subcommand for the action to perform. Currently, the only valid commands are \texttt{\%pack} and \texttt{\%meld}.

\texttt{\%urth} will return \texttt{\%\&} if given a valid command as input, otherwise it will return \texttt{[0 \%bail 0xfffffff9 \%urth-bad]}. No other output is emitted.

\subsection{\%peek}

The \texttt{\%peek} command is used to perform a namespace read request (a.k.a. scry) using Arvo's external peek interface \texttt{arm +22 in arvo.hoon}\footnote{\url{https://github.com/urbit/urbit/blob/develop/pkg/arvo/sys/arvo.hoon#L1774}}. The argument to \texttt{\%peek} is the \texttt{nom} input to \texttt{+peek} in \texttt{arvo.hoon} (\texttt{lyc} is auto-filled as \texttt{[~ ~]}, i.e., "request from self"). That is to say that the argument to \texttt{\%peek} must have type:
\begin{verbatim}
$+  each  path
$%  [%once vis=view syd=desk tyl=spur]
    [%beam vis=view bem=beam]
==
\end{verbatim}
Practically speaking, this means that the input will look like one of these three examples:
\begin{verbatim}
[%& p=path]
[%| p=[%once vis=view syd=desk tyl=spur]]
[%| p=[%beam vis=view bem=beam]]
\end{verbatim}
Where:
\begin{itemize}
  \item \texttt{path} is a \texttt{[view beam]}, with the \texttt{view} passed in as a \texttt{coin}
  \item \texttt{view} is the vane code for the scry, as well as an optional care, possibly appended to the vane (e.g. \texttt{\%j}, \texttt{\%gx}, etc.)
  \item \texttt{beam} is a \texttt{[beak spur]}
  \item \texttt{desk} is used to auto-generate a \texttt{beak}: \texttt{[our desk now]}
  \item \texttt{spur} is the scry endpoint for the agent or vane
\end{itemize}

The output of a \texttt{\%peek} command is \texttt{[\%peek (unit (unit scry-output))]}, where \texttt{~} means that the scry endpoint is invalid, and \texttt{[~ ~]} means that the scry resolved to nothing.

\subsection{\%peel}

\texttt{\%peel} attempts to emulate a scry-like namespace, like the one used by Arvo and accessed by \texttt{\%peek}. The argument to \texttt{\%peel} should be a path. Valid paths result in a non-null \texttt{unit} containing the result of the scry. Invalid paths result in null (i.e. \texttt{~}). The valid paths and the data they return are:

\begin{verbatim}
/help   (unit (list path))  Supported %peel paths
/live   (unit ~)            Pier health check; succeeds if pier is running
/khan   (unit ~)            Khan health check; succeeds if Khan vane is running
/info   (unit mass)         Pier info as a mass
/v      (unit @t)           Returns version of the Vere binary as a cord
/who    (unit @)            Returns the Azimuth identity of the ship as an atom
\end{verbatim}

Note that the pier info above is returned as a \texttt{mass} report, i.e. type \texttt{(pair cord (each * (list mass)))}. This is not the same as the \texttt{|mass} memory report. \texttt{/mass} is meant to be a valid \texttt{\%peel} path which returns the \texttt{|mass} memory report, but it is currently unimplemented.

\section{\%lick}

Altho also dealing with interprocess communication, Lick was designed for a very different scenario than Khan:  to allow external processes, in particular hardware drivers, to intercommunicate with Urbit.  (This breached the Earth/Mars divide.)  Thus \texttt{/sys/vane/lick} focuses on instrumenting a low-level noun interfaces over domain sockets.

\texttt{\%lick} manages IPC ports, and the communication between Urbit applications and POSIX applications via these ports. Other vanes and applications ask \texttt{\%lick} to open an IPC port, notify it when something is connected or disconnected, and transfer data between itself and the Unix application.

Lick works by opening a Unix socket for a particular process, which allows serialized IPC communications.  These involve a jammed noun so the receiving process needs to know how to communicate in nouns.

The IPC ports Lick creates are Unix domain sockets (\texttt{AF\_UNIX} address family) of the \texttt{SOCK\_STREAM} type.

The connexions are made via filepaths in \texttt{.urb/dev} of the pier.

The format is:

\begin{verbatim}
V.BBBB.JJJJ.JJJJ...
\end{verbatim}

\begin{itemize}
  \item \texttt{V} is the version
  \item \texttt{B} is the jam size in bytes (little endian)
  \item \texttt{J} is the jammed noun (little endian)
\end{itemize}

The process on the host OS must therefore strip the first 5 bytes, \texttt{++cue}\footnote{\url{https://docs.urbit.org/language/hoon/reference/stdlib/2p#cue}} the jamfile, check the mark and (most likely) convert the noun into a native data structure.

\subsection{\texttt{/sys/lull} Definition}

\begin{verbatim}
::                                              ::::
::::                    ++lick                    ::  (1j) IPC
  ::                                            ::::
++  lick  ^?
  |%
  +$  gift                                      ::  out result <-$
    $%  [%spin =name]                           ::  open an IPC port
        [%shut =name]                           ::  close an IPC port
        [%spit =name =mark =noun]               ::  spit a noun to the IPC port
        [%soak =name =mark =noun]               ::  soak a noun from the IPC port
    ==
  +$  task                                      ::  in request ->$
    $~  [%vega ~]                               ::
    $%  $>(%born vane-task)                     ::  new unix process
        $>(%trim vane-task)                     ::  trim state
        $>(%vega vane-task)                     ::  report upgrade
        [%spin =name]                           ::  open an IPC port
        [%shut =name]                           ::  close an IPC port
        [%spit =name =mark =noun]               ::  spit a noun to the IPC port
        [%soak =name =mark =noun]               ::  soak a noun from the IPC port
    ==
  ::
  +$  name  path
  --  ::lick
\end{verbatim}

To evaluate what \texttt{/sys/vane/lick} is doing, we need to look at Unix's IPC model briefly.  IPC ("interprocess communication") describes any way that two processes in an operating system's shared context have to communicate with each other.  \texttt{\%lick} focuses on Unix domain sockets\footnote{\url{https://en.wikipedia.org/wiki/Unix_domain_socket}}, which are just communication endpoints\footnote{\url{https://man7.org/linux/man-pages/man7/unix.7.html}}.

For instance, a valid use of \texttt{\%lick} would use cards that look like this:

\begin{verbatim}
++  init  [[%pass / %arvo %l %spin /control]~ this]
::
++  on-arvo
  |=  [=wire =sign-arvo]
  ?+  sign-arvo  (on-arvo:def wire sign-arvo)
      [%lick %soak *]
      ?+  mark.sign-arvo  [~ this]
      ::
        %connect
      ~&  >  "connect"
      :_  this  [%pass /spit %arvo %l %spit /control %init area.state]~
  ==  ==
::
++  send-state
  |=  =state
  ^-  card:agent:gall
  [%pass /spit %arvo %l %spit /control %state [slick:state face.state food.state live.state]]
\end{verbatim}

The vane definition of \texttt{/sys/vane/lick} is even simpler than \texttt{/sys/vane/khan}:  it has no \texttt{++abet} core and primarily communicates to the \texttt{unix-duct} in its state.  The \texttt{owner} is a \texttt{duct} to handle the return \texttt{\%soak}.

Gall needs to wrap \texttt{\%soak} and \texttt{\%spit} to route properly.  See e.g. \texttt{\%++ap-generic-take}.  This lets multiple agents share sockets with the same name, and each agent can have its own folder.

%\section{Background and Literature}

%Exposite the relevant background and literature.

%\begin{enumerate}
  %\item  Prefer to \texttt{enumerate}.
  %\item  It is easier to refer to than \texttt{itemize}.
%\end{enumerate}

%\noindent
%Mark \texttt{\textbackslash noindent} in paragraphs that continue the thought of an \texttt{\textbackslash enumerate} block.
%
%\section{Urbit's Implementation}
%
%Oftentimes you will then turn to exploring how Urbit or closely related systems like Nockchain have approached or considered a problem.
%
%\lstset{language=C}
%\begin{lstlisting}
%#include <stdio.h>
%
%double compute(float a, float b) {
    %return a * a + b * b;
%}
%
%int main() {
    %float x = 2.5;
    %float y = 3.7;
%
    %double result = compute(x, y);
%
    %printf("Result: %lf\n", result);
%
    %return 0;
%}
%\end{lstlisting}
%
%\noindent
%That code snippet was part of the text, and not a standalone entity to which we make separate repeated reference.
%
%\begin{figure}
  %\begin{lstlisting}[language=Python, caption={Example Python Code}, label={lst:example}]
  %def hello_world():
      %print("Hello, world!")
  %\end{lstlisting}
%\end{figure}
%
%Refer to the code listing: Listing~\ref{lst:example}.  (Note that we don't suppress indentation since it's a float.)
%

%\section{A Wild Ernest Appears}
%
%You can have as many sections as make sense.  Only sections and subsections appear in the table of contents.
%
\section{Conclusion}
